### 2020-12-03 paper/2020-10-revision-background.tex
Point commit before major outline changes

% Economic conditions
The dramatic improvements in economic conditions in India are best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive, people substitute 
away from that good and towards other goods.
With higher wages, this means that people work more and spend less time on non-wage 
earning activities, such as children or leisure.
The income effect captures that, holding hours worked constant, higher wages increase the 
available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

@@

% Son preference 
Empirically, son preference has only a minor effect on average fertility, but does impact 
the sex composition and family size within families through differential stopping behavior 
\citet{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and  
worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

@@

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest proxies in 
the data for wages and access to and cost of sex selection, none of which are available.
In addition, earlier research show that both higher female education and urbanization are 
associated with lower fertility and increased use of sex selection 
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 20 years or older, whether married or not, in the 
four rounds of the NFHS.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas.

@@

Families in urban areas are likely to have fewer children and higher rates of sex 
selection 
Two reasons fertility is lower in urban than rural areas are the higher cost of 
children and the lower mortality risk that comes from better access to health care.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
However, the better access to health care also means that there is easier access to
sex selection and the higher income of urban households makes it more affordable.

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but also increase the use of sex 
selection.

However, female education undoubtedly represents more than the value of women's time.
Beside propagation of low fertility norms, two examples are the productions of child 
health and offspring human capital.

First, children of more educated women tend to have better health outcomes and lower 
mortality than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Especially of interest here is that, although short birth spacing generally leads to worse 
health outcomes, the more educated the mother, the more muted this relationship is, which, 
at least partly, removes the health outcome incentive for longer birth spacing
\citep{Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
[need something on son preference? Maybe simply use increasing returns to male education]
Hence, families that value sons' education more will tend to both demand better-educated
women as mothers and have fewer children because both allow for higher quality boys.
With the increasing return to male education that India has experienced, this imply
an increased demand for better-educated women combined with lower fertility, and an 
increased use of sex selection when available.
In this case, the associations between higher female education, lower fertility, and more
sex selection are not a causal effect driven by higher female wages but instead the result 
of the increased return to male education.

Even as the level of female education has increased, the female labor force participation 
in both urban and rural areas has stagnated or decreased and is now one of 
the lowest in the world
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with women with either no education or 12 or more 
years the most likely to participate in the labor force and women with 8--11 years of 
education the least.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak to this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

% Sanskritization

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower-castes females gain access to 
education and their husbands' income increases, they adopt higher-caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.

>>>>

% Economic conditions

The dramatic improvements in economic conditions in India are best illustrated by the 
doubling of real wages between 1987 and 2011 \citep{Klasen2015}.
Economic theory predicts that the effect of higher wages on fertility is determined by 
the relative strengths of the substitution and income effects.
The substitute effect captures that as a good becomes more expensive, people substitute 
away from that good and towards other goods.
With higher wages, this means that people work more and spend less time on non-wage 
earning activities, such as children or leisure.
The income effect captures that, holding hours worked constant, higher wages increase the 
available income. 
This positive income effect leads people to spend more time on time-intensive activities 
such as children and leisure and less time working.

@@

% Son preference 
Empirically, son preference has only a minor effect on average fertility, but
differential stopping behavior does affect the sex composition and fertility of
individual families \citet{repetto72,leung94,clark00,Basu2010,Barcellos2014}. 
Furthermore, as mentioned above, stronger son preference leads to shorter spacing and  
worse health outcomes for daughters when there are no sons in the household 
\citep{Whitworth2002,Bhalotra2008,Maitra2008,Jayachandran2011,Jayachandran2017a}.

@@

% Factors/predictions
Female education and area of residence are the two principal explanatory variables I use
in the empirical analyses.
The main reason is that female education and area of residence are the closest 
available proxies wages and the access to and cost of sex selection, none of which are 
available in the data.
In addition, earlier research show that both higher female education and urbanization are 
associated with lower fertility and increased use of sex selection, and, therefore,
likely also changes in birth spacing
\citep{das_gupta97,dreze01,bhat03,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Finally, India has seen tremendous increases in both education and urbanization.
The proportion of the population living in urban areas almost doubled from 18\%
in 1960 to 35\% in 2019 \citep{United-Nations2019}.
Female education has increased substantially in both rural and urban areas,
but is still considerably higher in urban than in rural areas, as illustrated by Figure 
\ref{fig:education_over_time}, which shows the distribution of schooling by birth 
cohort for all urban and rural women, 20 years or older, using the four rounds of the NFHS.

@@

There are three reasons that families in urban areas are likely to have fewer children and 
higher rates of sex selection. 
First, the higher cost of children in urban than rural areas lowers fertility and 
increases the use of sex selection.
Second, better access to health care in urban areas lowers the mortality risk, further 
reducing fertility.%
\footnote{
However, holding parental education constant, there is evidence that children in
Indian slums do worse than they would have in rural areas \citep{Portner2018a}.
}
Finally, the better access to health care also imply easier access to sex selection and 
the higher income of urban households makes it more affordable.

Wages generally increase with education suggesting that higher female education
should shorten birth spacing and lower fertility, but the low and declining female labor 
force participation, especially for younger women, suggests that families face little 
incentive to space children more closely together for economic reasons.
Even as the level of female education increased, the female labor force participation 
in both urban and rural areas stagnated or decreased 
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.

Female labor force participation does, however, still depend on education.
In line with other countries, the NFHS data show a U-shaped relationship between 
education and working for married women, with women with either no education or 12 or more 
years the most likely to participate in the labor force and women with 8--11 years of 
education the least \citep{Goldin1994,Chatterjee2018}.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

One explanation for the low female labor force participation, especially for women
with 8--11 years of education, is that household income has increased so substantially 
that the income effect dominates any substitution effect.
Two findings support this interpretation.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

Furthermore, female education undoubtedly represents more than the value of women's time.
Beside propagation of low fertility norms, two examples are the productions of child 
health and offspring human capital.

First, children of more educated women have better health outcomes and lower mortality 
than less educated women, either from higher income or better health knowledge 
\citep{Rosenzweig1982a,Kovsted2002,Whitworth2002,Maitra2008}.
Especially of interest here is that, although short birth spacing generally leads to worse 
health outcomes, the more educated the mother, the more muted this relationship is, which, 
at least partly, removes the health outcome incentive for longer birth spacing
\citep{Molitoris2019}.

Second, better-educated women are more efficient at producing child human capital 
\citep{Behrman1999}.
Hence, the increasing return to male education may lead to an increased demand for 
better-educated mothers, which, combined with lower fertility and an increased use
of sex selection when available, allow families to achieve higher quality sons.
In this explanation, the associations between higher female education, lower fertility, 
and more sex selection are not a causal effect driven by higher female wages but instead 
the result of the increased return to male education.

% Sanskritization
Although the expansion in female education might bring in groups with initially different 
norms, the process of ``Sanskritization'' implies that as lower-castes females gain access 
to education, they adopt higher-caste norms such as stronger son preference and a retraction 
from the formal labor market \citep{Srinivas1956}.
The low and declining female labor force participation as husbands' wages increase 
is one indication that this process still operates \citep{Abraham2013,Chatterjee2018}.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Finally, even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.
